Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2023-03-01 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

Enclosed please find a major revision of manuscript number MBE-23-0015, titled ``Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory''. We are very grateful to you for co-ordinating the review process. We are also thankful to both the reviewers for their detailed analysis of our manuscript, which allowed us to improve our own understanding of our results, and helped us to improve our work. We were able to adddress the reviewers' concerns by performing additional analyses of our data, as well as by revising our manuscript.

Before embarking on a point-by-point reply to the reviewers' comments, we summarize the major structural changes we have made in our manuscript:

\begin{itemize}
\item We have revised our model based on several suggestions by the reviewers. We include the results from our older model in the supplementary material, for the reader's interest. 
\item We have changed figure 1 to a three panel figure such that it shows probabilities of fixation under different selection conditions.
\item In response to the reviewers' comments, we have also made several changes to the main text. We have highlighted these changes using a different font color.
\end{itemize}

Our manuscript is longer than the optimal 20 pages. This is primarily because we include a detailed methods section which we believe is essential for a reader to understand the results. Methods are especially important for theoretical papers that are not based on standard protocols and analytical pipeline. Thus we request you to allow us keep the methods in the main text and not in the supplementary material.

@@

We next focus on sequences that initiate transcription. Although several DNA sequence motifs corresponding to enhancers and core promoters have been identified, there is no set of sequence motifs that is deemed to be absolutely necessary for this process. The best understood, and the most common core promoter motifs are TATA-box and Inr, which we had included in our original analysis. We further note that although some sequences are indeed associated with enhancers, the actual functional enhancers are largely determined by epigenetic marks that vary significantly in different cell types, developmental stages, and during different environmental conditions. One common feature of both promoters and enhancers is that they both are characterized by a nucleosome depleted region (\href{https://doi.org/10.1038/s41576-019-0173-8}{Andersson and Sandelin, 2020}). Overall, we surmise that transcription can be initiated at any nucleosome depleted region but the stability of such transcripts is determined by polyA signals. We now explain these arguments in lines 95 -- 105.

We assume that such nucleosome free regions are uniformly distributed in open chromatin and these regions can initiate transcription at a basal rate. To this end, we first estimated the intergenic genome of \textit{D. melanogaster} that are in an open chromatin state (\href{https://doi.org/10.1186/1471-2164-15-988}{Milon et al., 2014}). Next, we calculated the fraction of known cis-regulatory elements (including enhancer) that are present in the open intergenic chromatin regions. This fraction is our estimate of basal rate of transcription initiation. Using this estimate, we now calculated the probability of transcript emergence based on poly-A signals alone, while ignoring the requirement of a core promoter. We first asked if such an assumption would significantly change the results presented in our first draft (Figures 1 -- 3). We found that these results do not change significantly. The fundamental inferences remain the same (for example, compare figures 1 and S1). We now mention the similarities and differences between the promoterless (new) and with-promoter (old) models in the main text (lines 234 -- 236, 276 -- 278, 308 -- 310). If we had data on every promoter/enhancer sequence, and if we incorporate them in the model, then the exact RNA probability values should lie in between those estimated by the promoterless model and TATA/Inr-only model. Therefore, in the revised manuscript we now present the results from the promoterless model, and provide a suitable justification in the text (lines 123 -- 126) including a new materials and methods section on how we inferred the general likelihood of transcription initiation (714 -- 733). We include the old results based on TATA/Inr-only model in the supplementary data (Fig S1 -- S3).

@@

We thank the reviewer for their suggestion and cite the relevant work in lines 470 -- 471.

{\itshape Lastly, I couldn't find the actual parameter values used for the various probabilities in the paper. I know the authors will make the code available, but for the casual reader a table would be valuable.}

Our primary parameters are mutation rate and mutation rate bias (shown in Table 1). All other parameters are derived and we therefore do not mention their values explicitly. Furthermore, the derived parameters are complex functions of gene length and nucleotide composition, and therefore, vary significantly. We now include the frequencies of trimers and hexamers from \textit{D. melanogaster} intergenic regions as supplementary files.

We will indeed make the code available when the work is published. Currently it is not accessible to the public because we use the same github repository to also keep track of the manuscript.

### 2023-03-01 Manuscripts/deNovoRates.tex
modified:   Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex 	modified:   Manuscripts/deNovoRates.tex

@@

\title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
\author{Bharat Ravi Iyengar$^1$, Erich Bornberg-Bauer$^{1,2}$}
\date{\small $^1$Institute for Evolution and Biodiversity, Westfalian Wilhelms -- University of M\"{u}nster, H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany\\ $^2$Max Planck-Institute for Biology T\"{u}bingen, T\"{u}bingen, Germany}

@@

\section{Introduction}

@@

\section{Results}

@@

\subsection{How likely is gene loss?}

@@

\subsection{Does \textit{de novo} gene emergence follow a preferred trajectory of events?}

@@

\subsection{Would extensive transcription loss suggest negative selection of toxic proteins?}

@@

\subsection{Does mutation bias shape protein composition?}

@@

\section{Discussion}

@@

\section{Materials and Methods}

@@

\section{Data availability}
We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution. Specifically, our model is implemented in the script \texttt{DeNovoProb.jl} which in turn uses the script \texttt{nucleotidefuncts.jl} for some basic functions.

>>>>

\renewcommand{\subsectionautorefname}{Methods section}
\renewcommand{\subsubsectionautorefname}{Methods section}
\renewcommand{\thesubsection}{\arabic{subsection}}

@@

\title{Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory}
\author{Bharat Ravi Iyengar$^{1,\dagger}$, Erich Bornberg-Bauer$^{1,2}$}
\date{\small $^1$Institute for Evolution and Biodiversity, University of M\"{u}nster,\\ H\"{u}fferstrasse 1, 48149 M\"{u}nster, Germany\\[1ex] $^2$Department of Protein Evolution, Max Planck Institute for Biology T\"{u}bingen, Max-Planck-Ring 5, 72076 T\"{u}bingen, Germany \\[1ex] $^\dagger$ Corresponding author: b.ravi@uni-muenster.de}

@@

\section*{Introduction}

@@

\section*{Results}

@@

\subsection*{How likely is gene loss?}

@@

\subsection*{Does \textit{de novo} gene emergence follow a preferred trajectory of events?}

@@

\subsection*{Would extensive transcription loss suggest negative selection of toxic proteins?}

@@

\subsection*{Does mutation bias shape protein composition?}

@@

\section*{Discussion}

@@

\section*{Materials and Methods}

@@

\section*{Data availability}
We performed all calculations using Julia programming language, and all scripts are freely available on GitHub: BharatRaviIyengar/DeNovoEvolution. Specifically, our model is implemented in the script \texttt{DeNovoProb.jl} which in turn uses the script \texttt{nucleotidefuncts.jl} for some basic functions.

\section*{Acknowledgments}

We thank Pete Czuppon (IEB, University of M\"{u}nster), for discussions on mathematical models.
